Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2016-04-28 curvature.tex
fixed usage of small
A proof EM can understand.

Now, we show that the average size of the moved subtrees is small.

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 4$ leaves.
	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 3}{4} .$$
	\label{lem:spr_size_mean}
\end{lem}
\begin{proof}
	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
	There are $n$ subtrees with 1 leaf, and 1 subtree each with $2,3, \ldots n - 1$ leaves.
	We ignore the root subtree, which does not induce any rSPR moves.
	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a 1-leaf subtree.
	We can bound the distribution of the remaining moves with the (loose) assumption that each other subtree contributes the same number of moves.
	In particular, we assume that each subtree contributes $2n - 3 - 5$ moves.
	This underestimates the number of small moved subtrees and overestimates the number of larger moved subtrees.
	Thus, this estimate bounds the mean moved subtree size from above.
	The mean subtree size, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.

@@

As we already showed in Lemmas~\ref{lem:spr_size} and~\ref{lem:spr_size_mean}, most adjacent trees differ by a small subtree.
This flatness-in-the-limit theorem thus helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.

>>>>

Now, we bound the average size of the moved subtrees.

\begin{lem}
	Let $T$ be a rooted tree with $n \ge 4$ leaves.
	Let $N_i(T)$ be the neighbors of $T$ induced by rSPR moves of an $i$-leaf subtree.
	Then $$\frac{\sum_i i \size{N_i(T)}}{\size{N(T)}} \le \frac{n + 3}{4} .$$
	\label{lem:spr_size_mean}
\end{lem}
\begin{proof}
	Again recall our distribution of rSPR moves from a tree in Lemma~\ref{lem:compute_degree}.
	By similar arguments as in the proof of Lemma~\ref{lem:spr_size}, a ladder tree has the greatest average moved subtree size (in terms of the number of leaves).
	There are $n$ subtrees with 1 leaf, and 1 subtree each with $2,3, \ldots n - 1$ leaves.
	We ignore the root subtree, which does not induce any rSPR moves.
	Thinking of the ratio as a weighted average of $i$, we can split the numerator and the denominator of the average into terms with $i \leq 2$ and those with $i > 2$.
    $N_i(T) \leq 2n - 3 - 5$ for the first set of terms, and $N_i(T) > 2n - 3 - 5$ for the second, so by replacing $N_i(T)$ in each case by $2n-3-5$ we increase the weight for $i \leq 2$ and decrease the weight on $i > 2$.
	Thus, by doing this replacement we obtain a upper bound on the mean moved subtree size.
	This bound, ignoring small subtrees, is $(3 + (n-2)) / 2 = (n + 1) / 2$.
	By Lemma~\ref{lem:spr_size}, at least half of the rSPR moves involve a small subtree.
	Thus, the average size of a moved subtree over all SPR operations from a given tree is at most the weighted average of $1 / 2 + ((n + 1) / 2) / 2$ = $(n + 1) / 4 + 1 / 2 = (n + 3) / 4$.

@@

As we already showed in Lemmas~\ref{lem:spr_size} and~\ref{lem:spr_size_mean}, most adjacent trees differ by a small subtree.
We next show that curvature is small in such a case for pairs of trees with a sufficiently large number of leaves.
Thus this ``flatness-in-the-limit'' theorem thus helps explain our empirical results in Section~\ref{sec:results:curvature} showing that the distribution of curvature pairs tends more towards 0 as $n$ increases.
